The illusory promise of the Aligned Rank Transform

### 2025-06-22 sandbox.qmd
many changes

Used for quick testing... eeee

>>>>

::: {.callout-important appearance="simple"}
## Under Review {.unnumbered}
This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
:::

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Useful libraries
library(crosstalk)
library(kableExtra)
library(gridExtra)
library(lmerTest)
library(tidyverse)
library(plotly)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Source code for reading data from experimental results and for plotting
source("dataReaders.R")
source("plotting.R")
source("effects_plot.R")
source("interactions_plot.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

::: {#fig-eta-squared}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("data/effect-sizes-lognormal.csv", sep=",", header=TRUE, strip.white=TRUE)
data$effectX2 <- factor(data$effectX2)

ylab = "X1 - partial eta squared"
xlab = "X2 - magnitude of main effect (a2)"

margins <- list(l = 50, r = 0, b = 60, t = 0, pad = 0)
cbPalette = c("#8888ff", "#888888", "#E69F00", "#009E73", "#FF5E00")
cbPalette2 = c("#8888ff88", "#88888888", "#E69F0088", "#009E7388", "#FF5E0088")

df <- data[data$effectX1 == 0,]

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor")
  )

fig

df <- data[data$effectX1 == 1,]
```
Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
:::

::: {#fig-eta-squared-2}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
df <- data[data$effectX1 == 1,]

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor")
  )

fig

df <- data[data$effectX1 == 1,]
```
Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
:::

### 2025-06-22 text/40-interpretingEffects.qmd
many changes

**Dealing with interpretation issues.** To avoid such interpretation issues, we focus --- unless otherwise stated --- on effects that apply monotonic transformations to population distributions. This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined). In other words, if a treatment increases the population mean, it will also increase the population median. We present an example in @fig-distributions. The figure shows two population distributions corresponding to the two intermediate levels of difficulty of our illustrative example (see @fig-example). We observe that the increased difficulty of the task translates both the population mean and the median to the right. In this case, we expect a statistical test to reject the null hypothesis, no matter whether it tests the population mean, the median, or the overall distribution shape.

@@

In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees. When parallel main effects are present, interpreting interactions can be challenging for all methods --- and ART offers no solution to this problem.

### Debate on the interpretation of interactions
We further discuss interpetation issues in the broader context of generalized liner models (GLMs).

@@

$$
E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x1 x2)
$${#eq-glm}

@@

@@

### Our approach {#approach}
<!-- Our analysis shows that inference errors are not simply due to the lack of robustness of a statistical procedure. In the case of interaction effects, errors will also emerge when the procedure makes inference on the wrong scale. 
As @Wagenmakers:2012 explain, *"the dependent variable reflects merely the output of a latent psychological process",* and unfortunately, *"in most experimental paradigms the exact relationship between unobserved process and observed behavior is unknown ..."*

Ideally, a statistical procedure should lead to conclusions that capture the true effects on the latent variable of interest. But as we discussed above, this might not be the case for the rank transformation methods that we study. For example, all four methods (*PAR*, *RNK*, *ART*, and *INT*) suggest that task difficulty interacts with use of technique on perceived performance because they disregard the fact that the observed data are simply projections on a discrete ordinal scale. Our goal is to understand how ART and the other methods deal with such problems. 
-->
The responses in a generalized linear model can follow arbitrary distributions. A key challenge is how to simulate such distributions while maintaining control over the parameters of the linear submodel. We address this by adopting a latent variable modeling approach, in which a normally distributed latent variable is transformed into the observed response variable.

**Latent variables.** We assume here is a single latent variable $\mathcal{Y}$ that is distinct from the variable $Y$ we observe. For example, the latent variable may represent a population's performance potential, working memory capacity, the perceived utility of a new technology, or quality of life. For convenience, we assume that this latent variable is linear, continuous, and normally distributed. This is a common assumption in latent variable modeling, e.g., in diffusion-decision models that predict response time and error in two-choice decision tasks [@Ratcliff_diffusion:2008], and in ordinal models [@Liddell:2018]. This assumption allows us to define null hypotheses within a shared, normalized space, regardless of the scale of the observed data.

@@

**Interpreting effects.** Our approach allows us to simulate main and interaction effects on the latent variable $\mathcal{Y}$ and observe how these effects manifest on the transformed variable $Y = \mathcal{T}(Y)$.

@@

2. The definition of the null hypothesis changes with the scale —- a null effect on the latent variable $\mathcal{Y}$ may not remain null when evaluated on the response variable $Y$.

For the first scenarios, we ensure that the interaction coefficient of the linear model is zero (e.g., $a_{12} = 0$ for the model in @eq-glm) when assessing main effects. When assessing interaction effects, we vary a single only coefficient that corresponds to a predictor (e.g., either $a_1$ or $a_2$), while keeping the others zero.

<!--
As discussed earlier, how to infer main effects is straightforward since we only study monotonic transformations here. Specifically, if we observe a main effect on the observed variable $Y'$, we can also conclude that there is a main effect on the latent variable $Y$. Notice, however, that if the observed variable is bounded or takes only discrete values, a main effect on $Y$ may not be observable.

The interpretation of interactions is more challenging. Depending on how the statistical procedure we use deals with different scales, observations of interactions among two or more factors on the transformed variable $Y'$ may lead to incorrect conclusions about the presence of interactions on the latent space of $Y$. Such errors emerge when the main effect of all interacting factors on the latent (or observed) variable is non-zero. In our analysis, we count them as statistical inference errors (Type I and II errors). Nevertheless, we try to distinguish them from typical inference errors, for which interaction interpretation issues do not come into place.
-->

>>>>

**Dealing with interpretation issues.** To avoid such interpretation issues, we focus -- unless otherwise stated -- on effects that apply monotonic transformations to population distributions. This also ensures a monotonic relationship between different measures of central tendency such as medians and means (with the exception of the Cauchy distribution, where the mean is undefined). In other words, if a treatment increases the population mean, it will also increase the population median. We present an example in @fig-distributions. The figure shows two population distributions corresponding to the two intermediate levels of difficulty of our illustrative example (see @fig-example). We observe that the increased difficulty of the task translates both the population mean and the median to the right. In this case, we expect a statistical test to reject the null hypothesis, no matter whether it tests the population mean, the median, or the overall distribution shape.

@@

In conclusion, focusing solely on the issues with the rank tranformation illustrated in @fig-interactions-rank is akin to missing the forest for the trees. When parallel main effects are present, interpreting interactions can be challenging for all methods -- and ART offers no solution to this problem.

### Controversy about the interpretation of interactions
We further discuss interpetation issues in the broader context of generalized liner models (GLMs).

@@

$$
E(Y|x_1, x_2) = g^{-1}(a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2)
$${#eq-glm}

@@

We emphasize that the existing literature does not clarify how ART is expected to interpret main and interaction effects in these scenarios. Early evaluations of ART did not consider nonlinear models, so the question was not even posed. @elkin:2021 who do study such models, use log-transformed values as their baseline for log-normal distributions (see their example on Page 4). So it seems that Elkin et al. assume ART's conclusions should align with the coefficients of the linear submodel. Unfortunately, their evaluations focus on contrasts and do not account for parallel effects, avoiding interpretation issues.

@@

### Our approach {#approach}
Responses in a generalized linear model can follow arbitrary distributions. A key challenge is how to simulate such distributions while maintaining control over the parameters of the linear submodel. We address this by adopting a latent variable modeling approach, in which a normally distributed latent variable is transformed into the observed response variable.

**Latent variables.** We assume there is a single latent variable $\mathcal{Y}$ that is distinct from the variable $Y$ we observe. For example, the latent variable may represent a population's performance potential, working memory capacity, the perceived utility of a new technology, or quality of life. For convenience, we assume that this latent variable is linear, continuous, and normally distributed. This is a common assumption in latent variable modeling, e.g., in diffusion-decision models that predict response time and error in two-choice decision tasks [@Ratcliff_diffusion:2008], and in ordinal models [@Liddell:2018]. This assumption allows us to define null hypotheses within a shared, normalized space, regardless of the scale of the observed data.

@@

**Interpreting effects.** Our approach allows us to simulate main and interaction effects on the latent variable $\mathcal{Y}$ and observe how these effects manifest on the transformed variable $Y = \mathcal{T}(\mathcal{Y})$.

@@

2. The definition of the null hypothesis changes with the scale: a null effect on the latent variable $\mathcal{Y}$ may not remain null when evaluated on the response variable $Y$.

Unless we explicitely mention interpretation issues, we focus on the first scenario. Note that in the case of discrete response variables, the function $\mathcal{T}$ is monotonic but not $strictly$ monotonic, since it involves a loss of information. Consequently, an effect on the latent variable may not be observable in the responses, or the observed magnitude of the effect may appear reduced.

### 2025-06-22 text/50-ExperimentalMethod.qmd
many changes

### Statistical modeling
To model the latent variable $Y$, we use a two-way (two factors) or a three-way (three factors) mixed-effects model. For simplicity, we explain here the model for two factors. Its extension to three factors is straightforward. The model has the following form:

@@

::: {#fig-effects}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
source("effects_plot.R")
plotEffects()
```
Varying $a_1$ or $a_2$ to control the magnitude of main effects. The plots show examples of population distributions on the latent variable $Y$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).    
:::

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\epsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y'$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.

@@

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $Y$. While partial $\eta^2$ is the most commonly used effect size measure, we chose to base our analysis on Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

@@

As we explained earlier, interaction effects are deformed when we transform the latent variable $Y$ to obtain the observed responses. Such transformations also affect the Type I error rates and power that we observe. Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $Y$. We make this distinction whenever it is relevant in our analysis.

>>>>

### Statistical modeling
To model the latent variable $\mathcal{Y}$, we use a two-way (two factors) or a three-way (three factors) mixed-effects model. For simplicity, we explain here the model for two factors. Its extension to three factors is straightforward. The model has the following form:

@@

::: {#fig-effects}
```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width = 9, warning=FALSE}
source("effects_plot.R")
plotEffects()
```
Varying $a_1$ or $a_2$ to control the magnitude of main effects. The plots show examples of population distributions on the latent variable $\mathcal{Y}$ for factors ($X_1$ or $X_2$) with two (top), three (middle), or four categorical levels (bottom).    
:::

We follow the approach of @DeBruine:2021 and use the R package *faux* v1.2.1 [@faux] to simulate data for our mixed-effects models. We assume that the distributions of random intercepts and errors are normal, or $s_k \sim N(0,\sigma_s)$ and $\epsilon_{ijk} \sim N(0,\sigma)$.
To simulate the observed variable $Y$, we then transform the response values $y_{ijk}$ as described in [Section 3](#approach). A key advantage of this method is that we can generate responses for any distribution, while we control effects on the latent variable in the exact same way.

@@

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$. While partial $\eta^2$ is the most commonly used effect size measure, we chose to base our analysis on Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

@@

As we explained earlier, interaction effects are deformed when we transform the latent variable $\mathcal{Y}$ to obtain the observed responses. Such transformations also affect the Type I error rates and power that we observe. Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $\mathcal{Y}$. We make this distinction whenever it is relevant in our analysis.

### 2025-06-22 text/60-Results-RatioScales.qmd
many changes

### Type I error rates in ratio scales {#ratio}
Our first experiments evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$. For the observed response variable $Y' = \mathcal{T}(Y)$, we evaluate transformations to four continuous and two discrete distributions:

1. No transformation, or $Y' = Y$. Distributions are normal with a grand mean $\mu = 0$ (e.g., see @fig-effects).

@@

**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases. The results show a very good behavior for RNK and INT across all distributions. The regular parametric ANOVA (PAR) keeps error rates below $5\%$. However, error rates become extremely low for some distributions, suggesting a loss in statistical power. We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021]. However, we also show that the method can be problematic with other non-normal distributions. As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels. This pattern is particularly pronounced under the log-normal distribution. We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero. In addition, error rates further grow when the sample size increases.

@@

@@

Finally, @fig-ratio-interaction-2 presents our results when the main effects on the two factors increase in parallel. Error rates become exceptionally high in some cases, growing up to $100\%$.

::: {#fig-ratio-interaction-2}
```{r, echo=FALSE, message=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readlyData(prefix, alpha, 0, distributions, dnames)

#xlab <- TeX("\\text{main effect of factors }X_1\\text{ and }X_2\\text{ }(a_1 = a_2)")
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
:::

However, these results require special attention since interaction interpretation issues come now into place. Let us examine the performance of each method:

- *PAR*. Error rates are high for all non-normal distributions, and become even higher when the sample size increases. As we explained in [Section 3](#interpretation), the parametric method cannot deal with scale differences that emerge when effects are transformed in a non-linear fashion. If we interpreted interactions based on absolute mean differences, disregarding their different scales (i.e., if we decided that the trend in @fig-interactions (left) is indeed a valid interaction effect), we would not observe these high error rates.

- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$). As shown in @fig-interactions-rank, the problem is due to the way the rank transformation deforms interactions.

- *INT.* The method exhibits a better behavior than RNK. Errors start again increasing for all distributions, but not until main effects become significantly large. An exception is the binomial distribution, for which the error rates of INT and RNK are similar. We also observe that for continuous distributions, both INT and RNK are scale invariant, since their performance is not affected by the choice of scale.

- *ART.* It keeps error rates at correct levels as long as population distributions are normal. For all other distributions, its error rates grow rapidly as effect sizes increase. We distinguish between two sources of errors: (i) lack of statistical robustness as observed in our previous tests; and (ii) interaction interpretation issues (see [Section 3](#interpretation)). Our results confirm that ART is not scale invariant. On the contrary, it is extremely sensitive to the scale of the observed data.

>>>>

### Type I error rates in ratio scales {#ratio}
We first evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$. For the observed response variable $Y = \mathcal{T}(\mathcal{Y})$, we evaluate transformations to four continuous and two discrete distributions:

1. No transformation, or $Y = \mathcal{Y}$. Distributions are normal with a grand mean $\mu = 0$ (e.g., see @fig-effects).

@@

**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.

@@

 The results show a very good behavior for RNK and INT across all distributions. The regular parametric ANOVA (PAR) keeps error rates below $5\%$. However, error rates become extremely low for some distributions, suggesting a loss in statistical power. We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021] but we also show that the method is problematic with other non-normal distributions. As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels. This pattern is particularly pronounced under the log-normal distribution. We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero. In addition, error rates further grow when the sample size increases.

@@

The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as those arise only when both coefficients $a_1$ and $a_2$ are non-zero. Since there can be different possible definitions of an interaction effect in these cases, we address them in a separate subsection.

### 2025-06-22 text/61-Results-OrdinalScales.qmd
many changes

### Type I error rates in ordinal scales {#ordinal}
Our second experiment evaluates Type I error rates for ordinal scales. We test again a $4 \times 3$ repeated-measures design. We focus on individual Likert items levels and implement an ordered-probit method [@Liddell:2018] to discretize the latent variable $Y$ into 5, 7, or 11 ordinal levels. To derive the discretization thresholds, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the responses $y_{ijk}$. We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range. @fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.

@@

**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect. Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales. In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases. Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.

@@

**Interaction effects**. @fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$. The observed patterns align with our previous findings; ART consistently inflates error rates, although to a lesser extent now.

@@

Finally, @fig-ordinal-interaction-2 presents our results when main effects are applied to both factors. None of the methods successfully maintains low error rates across all conditions. Intriguingly, we note that ART consistently performs worse than PAR. INT performs worse than ART and PAR in one specific case ($a_2, a_2 = 8$ in a scale with 11 equidistant levels), but overall, it exhibits a better behavior than all other methods.

::: {#fig-ordinal-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readlyData(prefix, alpha, 0, distributions, dnames)
plotlyError(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 105)

```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ 
:::

>>>>

### Type I error rates in ordinal scales {#ordinal}
We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ repeated-measures design. We focus on individual Likert items levels and implement an ordered-probit method [@Liddell:2018] to discretize the latent variable $Y$ into 5, 7, or 11 ordinal levels. To derive the discretization thresholds, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the responses $y_{ijk}$. We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range. @fig-ordinal presents examples of equidistant and flexible thresholds for a 5-level scale when the magnitude of main effect of $X_1$ is either $a_1 = 2$ or $a_1 = 8$, while all other effects are zero.

@@

**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.

@@

Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales. In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases. Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.

**Interaction effects**. @fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$. The observed patterns align with our previous findings; ART consistently inflates error rates.

@@

**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger. As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null.

::: {#fig-ordinal-large-samples}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "2_test_4x3_Ordinal_sample_size"

distributions <- c("likert5", "likert5B", "likert7", "likert7B", "likert11", "likert11B")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")

df <- readlyData1(prefix, alpha, 0, distributions, dnames)
plotlyErrorByEffect(df, xlab = "n", max = 100)
```
Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
:::

We will observe similar trends if we test other discrete distributions, such as binomial distributions. These results confirm that ART is inappropriate for discrete data.
